On the uniqueness and stability of dictionaries for sparse representation of noisy signals

### 2017-04-07 manuscripts/PNAS/chazthm_pnas.tex
chaz SR

\significancestatement{Many naturally occurring signals (visual scenery, speech, electrocardiograms, etc.) can be characterized as parsimonious combinations of elementary waveforms drawn from a large dictionary. Absent analytic descriptions of these signal classes, it falls on algorithms from machine learning to extract such structures given many examples from a class. We give conditions guaranteeing when a dictionary of a given size is uniquely and stably determined by data, regardless of the algorithm used to compute it. These results provide theoretical grounding for  the application of dictionary learning to recovering sparse signals from unknown noisy compressive measurements.}

@@

% If your first paragraph (i.e. with the \dropcap) contains a list environment (quote, quotation, theorem, definition, enumerate, itemize...), the line after the list may have some extra indentation. If this is the case, add \parshape=0 to the end of the list environment.
\dropcap{B}lind source identification is a classical problem in signal processing \cite{sato1975method}.  
One popular modern formulation of the idea is to find a dictionary of $m$ elementary waveforms, at most $k$ of which need be linearly combined to represent each $n$-dimensional signal in a dataset of size $N$, typically where $k < m \ll N$.  Approximating solutions to this problem have
provided insight into the structure of many signals lacking explicit analytic structure (see \cite{Zhang15} for a comprehensive review). 
In particular, it has been shown that response properties of simple-cell neurons in mammalian visual cortex emerge from learning a dictionary for sparse linear coding of natural images \cite{Olshausen96, hurri1996image, bell1997independent, van1998independent}, a major advance in computational neuroscience. A fundamental aspect of this finding is that the latent waveforms (``Gabors'') estimated from data appear to be canonical; i.e., they are found in learned dictionaries independent of algorithmic implementation or natural image training set.

@@

Motivated by these discoveries and earlier work in the theory of neural communication \cite{Coulter10, Isely10}, we address when dictionaries and sparse representations are uniquely determined by data.  Answers to this question also have other real-world implications.  For example, a sparse coding analysis of local painting style can be used for forgery detection \cite{hughes2010, Olshausen10}, but only if sparse components of the artwork are independent of implementation idiosyncrasies. Fortunately, algorithms with proven recovery of generating dictionaries under certain 
conditions have recently been proposed (see \cite[Sec.~I-E]{Sun16} for a summary of the state-of-the-art).  However, there are few universal theorems explaining the general phenomenon of uniqueness.

Here, we prove very generally that uniqueness in sparse linear coding is an expected property of the model.  In other words, whenever enough sparsely represented data generated from a dictionary (satisfying a spark condition) are observed, the original codes and dictionary are uniquely determined up to an error that is commensurate with the measurement noise.  Applications of this result include:  1) a sufficient diversity of sparse codes sent through a random linear compressive channel are unique, 2) the sparse linear coding problem is well-posed in the sense of Hadamard \cite{Hadamard1902}, and 3) an explanation for universal representation of natural signals in neuroscience.

@@

Our main technical finding is that a sufficient quantity of measurements generated from any dictionary satisfying the spark condition 
has a stable sparse representation.  More specifically, such dictionaries are uniquely identifiable from as few as \mbox{$N = m(k-1){m \choose k} + m$} sparse linear combinations of their columns up to an error that is linear in the measurement noise (Thm.~\ref{DeterministicUniquenessTheorem}). In fact, provided $n \geq k$, %$(n, m, k)$ satisfy CS inequality \eqref{CScondition}, 
in almost all cases the dictionary learning problem is well-posed given enough data (Cor.~\ref{ProbabilisticCor}). Importantly, these guarantees hold without assuming the recovered matrix satisfies the spark condition. Our results  also apply when only an upper bound on the number of dictionary elements is known. The explicit, algorithm-independent criteria we provide should be a useful tool in the theory of sparse dictionary learning.

In the next section, we give precise statements of our main results, which include an explicit form for $\varepsilon(\delta_1, \delta_2)$. We then prove our main theorem (Thm.~\ref{DeterministicUniquenessTheorem}) in Sec.~\ref{DUT} after stating some additional definitions and lemmas required for the proof, including our main tool from combinatorial matrix analysis (Lem.~\ref{MainLemma}). All other proofs are relegated to the Supplemental. 
Finally, we present several applications in Discussion Sec.~\ref{Discussion}.

@@

There are other less direct consequences of Thm.~\ref{DeterministicUniquenessTheorem} to stability in sparse linear coding.  The following implication relates the result to more common optimization formulations.

\begin{corollary}\label{SLCopt}
If the assumptions of Thm.\ref{DeterministicUniquenessTheorem} hold with more than $(k-1)\left[ {\bar m \choose k} + |E|k{\bar m \choose k-1}\right]$ vectors $\mathbf{x}_i$ supported on each $S \in E$, then all matrices $\mathbf{B} \in \mathbb{R}^{n \times \bar m}$ and $\mathbf{\bar x}_1, \ldots, \mathbf{\bar x}_N \in \mathbb{R}^{\bar m}$ solving:
\begin{align}\label{minsum}
\min \sum_{i = 1}^N \|\mathbf{\bar x}_i\|_0 \ \ 
\text{subject to} \ \ \|\mathbf{A}\mathbf{x}_i - \mathbf{B}\mathbf{\bar x}_i\|_2 \leq \varepsilon, \ \text{for all $i$},
\end{align}
%
necessarily satisfy implications 1) and 2) from Thm.~\ref{DeterministicUniquenessTheorem}.
%must have $\bar m \geq m$ and, provided $p = \bar m - \ell(\bar m - m) > 0$, necessarily satisfy \eqref{Cstable}
%%\begin{align}\label{Cstable}
%%\|(\mathbf{A}_J- \mathbf{B}_{\bar J} \mathbf{PD} )_j\|_2 \leq C_1 \varepsilon, \ \  \text{for  $j \in J$},
%%\end{align}
%for some $J \in {[m] \choose p}$, $\bar J \in {[\bar m] \choose p}$, permutation matrix $\mathbf{P}$, and invertible diagonal matrix $\mathbf{D}$.
\end{corollary}

@@

\section{Proof of Theorem~\ref{DeterministicUniquenessTheorem} and Corollary~\ref{SLCopt}}\label{DUT}

@@

In this note, we generalize the approach of \cite{Hillar15} to prove uniqueness in Prob.~\ref{InverseProblem} for the case of noisy measurements while also reducing the sufficient number of samples.
% FRITZ: Inverse problems instead of Data Analysis. Don't assume that everyone is assuming they are recovering ground truth. Cite results again and state implications! Focus on them (probabilistic ones too).
Our results provide theoretical grounding for the use of sparse linear coding in blind source separation, wherein the goal is to infer the generating dictionary and sparse codes from noisy measurements, \eqref{LinearModel}.  We also collect a set of useful mathematical tools and basic facts for future research.
The main motivation for this work, however, was to understand how seemingly universal and stable representations are discovered by training a sparse linear coding model on natural data, independent of specific algorithmic implementation.    We elaborate more on these applications below after discussing  theoretical aspects.  % We close with several application areas.

@@

What we have shown here is that as an inverse problem to be approximated by a numerical algorithm, the sparse linear coding model  generally produces \textit{well-posed} optimizations.  This early concept of Hadamard \cite{Hadamard1902} can be paraphrased as the idea that good parameterized models should pose optimization problems with solutions that are \textit{stable}; that is, continuous in the data.  In other words, perturbing a dataset by a small amount should only slightly influence the identifiability or uniqueness of parameters in the model, rather than discontinuously affect the character of optimization involved.  In this regard, we demonstrate a linear relationship, \eqref{epsdel}, between noise in the data and the unique solution in  sparse linear coding, with explicit constants expressed in terms of original latent parameters. 
Nonetheless, it would be of practical utility to determine the best possible dependence of $\varepsilon$ on $\delta_1, \delta_2$ in Def.\ref{maindef} as well as the minimal requirements on the number and diversity of generating codes, and we hope that other researchers continue to improve and extend our results.
% , although they are too computationally expensive to be of practical value.  We hope .

One notable component of our contribution is a combinatorial criteria (regular hypergraphs satisfying the singular intersection property, Def.~\ref{sip}) for choosing support sets for original sparse codes. Fully understanding those combinatorial designs allowing for stable representations of datasets is an interesting research area for the future.  For instance, whether there is a support set size $N$ that is polynomial in $m,k$ gauanteeing the conclusions of Thm.~\ref{DeterministicUniquenessTheorem} has implications for the computational complexity of Prob.\ref{InverseProblem} and related questions \cite{Tillmann15}.  We also remark that our main uniqueness result accounts for deterministic ``worst-case'' noise, whereas the ``effective'' noise might be much smaller when it is sampled from a given distribution; in such cases, the constants  %Thms.~\ref{DeterministicUniquenessTheorem},~\ref{DeterministicUniquenessTheorem2} 
will improve. %We note also that these results extend trivially to cases where point-wise injective nonlinearities are applied to the data.

A technical difficulty in our main theorem is the absence of a spark condition assumption on alternate sparse data factorizations.
Although mathematically interesting that no such requirement is necessary, there are other reasons to seek out such a theoretical guarantee. For instance, it is difficult to ensure an algorithm maintain a dictionary satisfying \eqref{SparkCondition} at each learning stage; indeed, even certifying  a dictionary has this property is likely intractable given its NP-hardness \cite{tillmann2014computational}.

>>>>

\significancestatement{Many naturally occurring signals (visual scenery, speech, electrocardiograms, etc.) can be characterized as parsimonious combinations of elementary waveforms drawn from a large dictionary. Absent analytic descriptions of these signal classes, it falls on algorithms from machine learning to extract such structures given many examples from a class. We give conditions guaranteeing when a dictionary of a given size is uniquely and stably determined by data, regardless of the algorithm used to compute it. We propose an explanation for universal representation of natural signals in sparse coding models from neuroscience.
%These results provide theoretical grounding for  the application of dictionary learning to recovering sparse signals from unknown noisy compressive measurements.
}

@@

% If your first paragraph (i.e. with the \dropcap) contains a list environment (quote, quotation, theorem, definition, enumerate, itemize...), the line after the list may have some extra indentation. If this is the case, add \parshape=0 to the end of the list environment.
\dropcap{B}lind source separation is a classic inverse problem in signal processing \cite{sato1975method}.
One constraint commonly imposed is that each of $N$ observed $n$-dimensional signals consists of a linear combination of at most $k$ elementary waveforms drawn from some unknown `dictionary' of size $m$, typically where $k < m \ll N$ (see \cite{Zhang15} for a comprehensive review of this and related models).
%A popular underlying assumption is that there exists a `dictionary' of $m$ elementary waveforms, at most $k$ of which need be linearly combined to represent each signal in a given dataset of size $N$, typically where $k < m \ll N$.  
Approximate solutions to this problem have provided insight into the structure of many signal classes lacking explicit analytic descriptions [cite?]. 
In particular, it has been shown that response properties of simple-cell neurons in mammalian visual cortex emerge from the optimization of a dictionary to represent small patches of natural images \cite{Olshausen96, hurri1996image, bell1997independent, van1998independent}, a major advance in computational neuroscience. A curious aspect of this finding is that the latent waveforms (``Gabors'') estimated from data appear to be canonical [cite?]; %\cite{donoho2001can}
i.e., they are found in learned dictionaries independent of algorithmic implementation or natural image training set.

@@

Motivated by these discoveries and earlier work in the theory of neural communication \cite{Coulter10, Isely10}, we address when dictionaries and sparse representations are uniquely determined by data.  Answers to this question also have other real-world implications.  For example, a sparse coding analysis of local painting style can be used for forgery detection \cite{hughes2010, Olshausen10}, but only if sparse components of the artwork can be consistently estimated. % are independent of implementation idiosyncrasies. 
Fortunately, algorithms with proven recovery of generating dictionaries under certain 
conditions have recently been proposed (see \cite[Sec.~I-E]{Sun16} for a summary of the state-of-the-art). There are few theorems one can cite to explain the general phenomenon of uniqueness, however.

Here, we prove very generally that uniqueness and stability in sparse linear coding is an expected property of the model. %; that is, it is a well-posed problem (in the sense of Hadamard \cite{Hadamard1902}). 
%In other words, whenever enough sparsely represented data generated from a dictionary (satisfying a spark condition) are observed, the original codes and dictionary are uniquely determined up to an error that is commensurate with the measurement noise. 
More specifically, dictionaries are identifiable from as few as \mbox{$N = m(k-1){m \choose k} + m$} sparse linear combinations of their columns up to an error that is linear in the measurement noise (Thm.~\ref{DeterministicUniquenessTheorem}). In fact, provided $n \geq k$, in almost all cases the dictionary learning problem is well-posed (in the sense of Hadamard \cite{Hadamard1902}) given enough data (Cor.~\ref{ProbabilisticCor}). Importantly, these guarantees hold without assuming the recovered matrix satisfies a spark condition. Our results  also apply when only an upper bound on the number $m$ of dictionary elements is known. The explicit, algorithm-independent criteria we provide should be a useful tool in the theory of dictionary learning.

Applications of our results include:  1) a sufficient diversity of sparse codes sent through a random linear compressive channel are identifiable and 2) an explanation for universal representation of natural signals in sparse coding models from neuroscience.

@@

%Our main technical finding is that a sufficient quantity of measurements generated from any dictionary satisfying the spark condition has a stable sparse representation.  More specifically, such dictionaries are identifiable from as few as \mbox{$N = m(k-1){m \choose k} + m$} sparse linear combinations of their columns up to an error that is linear in the measurement noise (Thm.~\ref{DeterministicUniquenessTheorem}). In fact, provided $n \geq k$, in almost all cases the dictionary learning problem is well-posed (in the sense of Hadamard \cite{Hadamard1902}) given enough data (Cor.~\ref{ProbabilisticCor}). Importantly, these guarantees hold without assuming the recovered matrix satisfies a spark condition. Our results  also apply when only an upper bound on the number of dictionary elements is known. The explicit, algorithm-independent criteria we provide should be a useful tool in the theory of dictionary learning.

In the next section, we give precise statements of our main results, which include an explicit form for $\varepsilon(\delta_1, \delta_2)$. We then prove our main theorem (Thm.~\ref{DeterministicUniquenessTheorem}) in Sec.~\ref{DUT} after stating some additional definitions and lemmas required for the proof, including our main tool from combinatorial matrix analysis (Lem.~\ref{MainLemma}). We also give a simple argument extending our uniqueness guarantees to the following more common optimization formulation of the sparse linear coding problem (Cor.~\ref{SLCopt}):

\begin{problem}\label{OptimizationProblem}
Find a matrix $\mathbf{B} \in \mathbb{R}^{\bar m}$ and $k$-sparse $\mathbf{\bar x}_1, \ldots, \mathbf{\bar x}_N \in \mathbb{R}^m$ that solve:
\begin{align}\label{minsum}
\min \sum_{i = 1}^N \|\mathbf{\bar x}_i\|_0 \ \ 
\text{subject to} \ \ \|\mathbf{A}\mathbf{x}_i - \mathbf{B}\mathbf{\bar x}_i\|_2 \leq \varepsilon, \ \text{for all $i$}.
\end{align}
\end{problem}

All other proofs are relegated to the Supplemental. 
Finally, we present several applications in Discussion Sec.~\ref{Discussion}.

@@

There are other less direct consequences of Thm.~\ref{DeterministicUniquenessTheorem} to stability in sparse linear coding. Consider the optimization problem posed in Prob.~\ref{SLCopt}.

\begin{corollary}\label{SLCopt}
If the assumptions of Thm.\ref{DeterministicUniquenessTheorem} hold with more than $(k-1)\left[ {\bar m \choose k} + |E|k{\bar m \choose k-1}\right]$ vectors $\mathbf{x}_i$ supported on each $S \in E$, then all solutions to Prob.~\ref{OptimizationProblem} necessarily satisfy implications 1) and 2) from Thm.~\ref{DeterministicUniquenessTheorem}.
\end{corollary}
%\begin{corollary}\label{SLCopt}
%If the assumptions of Thm.\ref{DeterministicUniquenessTheorem} hold with more than $(k-1)\left[ {\bar m \choose k} + |E|k{\bar m \choose k-1}\right]$ vectors $\mathbf{x}_i$ supported on each $S \in E$, then all matrices $\mathbf{B} \in \mathbb{R}^{n \times \bar m}$ and $\mathbf{\bar x}_1, \ldots, \mathbf{\bar x}_N \in \mathbb{R}^{\bar m}$ solving:
%\begin{align}\label{minsum}
%\min \sum_{i = 1}^N \|\mathbf{\bar x}_i\|_0 \ \ 
%\text{subject to} \ \ \|\mathbf{A}\mathbf{x}_i - \mathbf{B}\mathbf{\bar x}_i\|_2 \leq \varepsilon, \ \text{for all $i$},
%\end{align}
%
%necessarily satisfy implications 1) and 2) from Thm.~\ref{DeterministicUniquenessTheorem}.

%must have $\bar m \geq m$ and, provided $p = \bar m - \ell(\bar m - m) > 0$, necessarily satisfy \eqref{Cstable}
%%\begin{align}\label{Cstable}
%%\|(\mathbf{A}_J- \mathbf{B}_{\bar J} \mathbf{PD} )_j\|_2 \leq C_1 \varepsilon, \ \  \text{for  $j \in J$},
%%\end{align}
%for some $J \in {[m] \choose p}$, $\bar J \in {[\bar m] \choose p}$, permutation matrix $\mathbf{P}$, and invertible diagonal matrix $\mathbf{D}$.
%\end{corollary}

@@

\section{Proofs of Theorem~\ref{DeterministicUniquenessTheorem} and Corollary~\ref{SLCopt}}\label{DUT}

@@

In this note, we generalize the approach of \cite{Hillar15} to prove the stability of unique solutions to Probs.~\ref{InverseProblem} and \ref{OptimizationProblem} while also significantly reducing the known sample complexity.
% FRITZ: Inverse problems instead of Data Analysis. Don't assume that everyone is assuming they are recovering ground truth. Cite results again and state implications! Focus on them (probabilistic ones too).
Our results provide theoretical support to the application of the sparse linear coding model to blind source separation problems, wherein the goal is to infer the generating dictionary and sparse codes from noisy measurements, \eqref{LinearModel}. We also collect a set of useful mathematical tools and basic facts for future research.
The main motivation for this work, however, was to understand how seemingly universal and stable representations emerge from sparse coding models fit to natural data by a variety of methodologies. We elaborate more on these applications below after discussing theoretical aspects.  % We close with several application areas.

@@

What we have shown here is that the sparse linear coding model generally produces a \textit{well-posed} inverse problem to be approximately solved by a numerical algorithm. This early concept of Hadamard \cite{Hadamard1902} can be paraphrased as the idea that inferences from observations should be robust to the inevitable uncertainty in measurement; that is, continuous in the data. In other words, perturbing a dataset by a small amount should only slightly influence the inferred parameters. In this regard, for the sparse linear coding model we demonstrate a linear relationship, \eqref{epsdel}, between measurement noise and the deviation of any solution from the true parameters, with explicit constants expressed in terms of these parameters. Moreover, we show that even if the meta-parameter defining the number of dictionary elements is overestimated, a subset of parameters are still identifiable.
It would be of practical utility to determine the best possible dependence of $\varepsilon$ on $\delta_1, \delta_2$ in Def.\ref{maindef} as well as the minimal requirements on the number and diversity of generating codes, and we hope that other researchers continue to improve and extend our results.

%We note also that these results extend trivially to cases where point-wise injective nonlinearities are applied to the data.

One notable component of our contribution is a combinatorial criteria (regular hypergraphs satisfying the singular intersection property, Def.~\ref{sip}) for the support sets of sparse codes key to the identification of the dictionary. Fully understanding those combinatorial designs allowing for stable sparse representations is an interesting research area for the future. For instance, whether there is a support set size $N$ that is polynomial in $m,k$ gauranteeing the conclusions of Thm.~\ref{DeterministicUniquenessTheorem} has implications for the computational complexity of Probs.~\ref{InverseProblem} and \ref{OptimizationProblem} and related questions \cite{Tillmann15}. We also remark that our main uniqueness result accounts for deterministic ``worst-case'' noise, whereas the ``effective'' noise might be much smaller when it is sampled from a given distribution; in such cases, the constants  %Thms.~\ref{DeterministicUniquenessTheorem},~\ref{DeterministicUniquenessTheorem2} 
will improve.

A technical difficulty in proving Thm.~\ref{DeterministicUniquenessTheorem} was the absence of a spark condition assumption on solutions to Prob.~\ref{InverseProblem}.
Although mathematically interesting that no such requirement is necessary, there are other reasons to seek out such a theoretical guarantee. For instance, it is difficult to ensure that an algorithm maintain a dictionary satisfying \eqref{SparkCondition} at each iteration; indeed, even certifying a dictionary has this property is likely intractable given its NP-hardness \cite{tillmann2014computational}.
